# GEO200CN, Spring 2019

## Introduction

In this lab, you practice some basic R skills --- if you already know R, it should be a breeze. If you have not used R (much), carefully read the introductory material as suggested. The data files used in this lab are on the Canvas website.

__Hand in the results as a R Markdown file.__ Both the source file and a compiled html file. You can complete the .Rmd file available on Canvas.


## Part 1


__Carefully read chapter 1 to 6 in http://rspatial.org/intr/ __


__1) Read and explore data __

There are two text files with data (loblolly.dat) and (ChickWeightSub.csv). 

Write a script (including brief comments) that

a) Reads them into R
b) for both data sets, lists all the variable names
c) for each variable in one of the data sets, use a function to compute the min and a function to compute the mean value 
d) now pick one variable from one data set and use functions summary, table, quantile, and boxplot


## Part 2

__Carefully read chapter 7 to 14 in http://rspatial.org/intr/ __


__2) Functions__

a) Use the files from lab 1 to compute mean values for each variable using 'apply'

b) Write a function 'f' that behaves like this:

```{r eval=FALSE}
> f('Jim')
[1] "hello Jim, how are you"
```

c) Write a function 'sumOfSquares' that behaves like this:

```{r eval=FALSE}
> d <- c(1,5,2,4,6,2,4,5)
> sumOfSquares(d)
[1] 21.875
```

To compute the "sum of squares", subtract the mean value of all numbers from each number. Square these numbers and sum them


__3) Data preparation__. Use the data in "students.csv" and "scores.dat" to assess if there is a relation between the "group" or "age" of a student and their average score and for the three subjects (math, english, geography)

a) use reshape and merge to combine the two data sets
b) compute correlation coefficients
c) produce a relevant plot


