---
title: "GEO 200CN Lab 9"
author: "Kenneth Larrieu"
date: "May 29, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Spatial Distribution Models

## Data

### Observations

```{r}
# load bigfoot observation data
if (!require("rspatial")) devtools::install_github('rspatial/rspatial')
library(rspatial)
bf <- sp_data('bigfoot')
dim(bf)
head(bf)
```


```{r}
# plot bigfoot sightings
plot(bf[,1:2], cex=0.5, col='red')
library(maptools)
data(wrld_simpl)
plot(wrld_simpl, add=TRUE)
```

### Predictors

```{r}
library(raster)
# load climate data to use as predictor variables
wc <- raster::getData('worldclim', res=10, var='bio')
plot(wc[[c(1, 12)]], nr=2)
```

```{r}
# extract climate data at bigfoot sighting locations
bfc <- extract(wc, bf[,1:2])
head(bfc)
```

```{r}
# check for missing climate data
i <- which(is.na(bfc[,1]))
i
plot(bf[,1:2], cex=0.5, col='blue')
plot(wrld_simpl, add=TRUE)
# plot locations with missing climate data
points(bf[i, ], pch=20, cex=3, col='red')
```

```{r}
# plot temp vs precip for bigfoot locations
plot(bfc[ ,'bio1'] / 10, bfc[, 'bio12'], xlab='Annual mean temperature (C)',
         ylab='Annual precipitation (mm)')
```

### Background Data

```{r}
library(dismo)
# extent of all points
e <- extent(SpatialPoints(bf[, 1:2]))
e

set.seed(0)
# create random sampling of points
bg <- sampleRandom(wc, 5000, ext=e)
dim(bg)

head(bg)
```

Combining presence and background data:

```{r}
d <- rbind(cbind(pa=1, bfc), cbind(pa=0, bg))
d <- data.frame(d)
dim(d)
```

## Fit a Model

```{r}
# split into east and west data
de <- d[bf[,1] > -102, ]
dw <- d[bf[,1] <= -102, ]
```

### CART

```{r}
library(rpart)
# fit cart model for western bigfoot sightings: presence/absence as function of climate
cart <- rpart(pa~., data=dw)
printcp(cart)

plotcp(cart)

plot(cart, uniform=TRUE, main="Regression Tree", margin=0.2)
# text(cart, use.n=TRUE, all=TRUE, cex=.8)
text(cart, cex=.8, digits=1)
```


**Question 1:** Describe the conditions under which you have the highest probability of finding our beloved species?

Using the list of bioclimatic variable names we can describe the conditions where we are most likely to observe Bigfoot, according to the classification tree:

BIO1 = Annual Mean Temperature
BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp))
BIO3 = Isothermality (BIO2/BIO7) (* 100)
BIO4 = Temperature Seasonality (standard deviation *100)
BIO5 = Max Temperature of Warmest Month
BIO6 = Min Temperature of Coldest Month
BIO7 = Temperature Annual Range (BIO5-BIO6)
BIO8 = Mean Temperature of Wettest Quarter
BIO9 = Mean Temperature of Driest Quarter
BIO10 = Mean Temperature of Warmest Quarter
BIO11 = Mean Temperature of Coldest Quarter
BIO12 = Annual Precipitation
BIO13 = Precipitation of Wettest Month
BIO14 = Precipitation of Driest Month
BIO15 = Precipitation Seasonality (Coefficient of Variation)
BIO16 = Precipitation of Wettest Quarter
BIO17 = Precipitation of Driest Quarter
BIO18 = Precipitation of Warmest Quarter
BIO19 = Precipitation of Coldest Quarter

According to the classification tree, Bigfoots are likely to occur in regions with high temperature seasonality, a high mean temperature in the warmest quarter, and either i) low precipitation seasonality and high mean temperature in the wettest quarter, or ii) high precipitation seasonality high precipitation in the driest month, high max temperature in the warmest month, high isothermality, and low minimum temperature of the coldest month.

### Random Forest

```{r}
library(randomForest)
fpa <- as.factor(dw[, 'pa'])

crf <- randomForest(dw[, 2:ncol(dw)], fpa)
crf

plot(crf)
```

```{r}
# variable importance plot
varImpPlot(crf)
```

```{r}
trf <- tuneRF(dw[, 2:ncol(dw)], dw[, 'pa'])
trf
mt <- trf[which.min(trf[,2]), 1]
mt
```

**Question 2:** What did tuneRF help us find? What does the values of mt represent?

`tuneRF` Found the optimal value of the `mtry` parameter for the `randomForest` algorithm, which is the number of variables used as candidates for each node split. `mt` is the value of `mtry` which minimizes the OOB error; i.e. it is the value which produces the best random forest.

```{r}
rrf <- randomForest(dw[, 2:ncol(d)], dw[, 'pa'], mtry=mt)
rrf
plot(rrf)
```

```{r}
varImpPlot(rrf)
```

## Predict

```{r}
# Extent of the western points
ew <- extent(SpatialPoints(bf[bf[,1] <= -102, 1:2]))
ew
```

### Regression

```{r}
rp <- predict(wc, rrf, ext=ew)
plot(rp)
```

```{r}
eva <- evaluate(dw[dw$pa==1, ], dw[dw$pa==0, ], rrf)
eva
```

```{r}
plot(eva, 'ROC')
```

```{r}
# get threshold for classification
tr <- threshold(eva)
tr
# plot values above threshold
plot(rp > tr[1, 'spec_sens'])
```

### Classification

```{r}
rc <- predict(wc, crf, ext=ew)
plot(rc)
```

```{r}
rc2 <- predict(wc, crf, ext=ew, type='prob', index=2)
plot(rc2)
```

## Extrapolation

```{r}
de <- na.omit(de)
eva2 <- evaluate(de[de$pa==1, ], de[de$pa==0, ], rrf)
eva2
plot(eva2, 'ROC')
```

```{r}
eus <- extent(SpatialPoints(bf[, 1:2]))
eus
rcusa <- predict(wc, rrf, ext=eus)
plot(rcusa)
points(bf[,1:2], cex=.25)
```

**Question 3:** Why would it be that the model does not extrapolate well?

Our model might not exrtapolate well because the sightings in the east may be a different species than the western Bigfoot, with significantly different climatic preferences. It appears that the reported observations in the east often occur in very different climates than those of the western observations, which supports this hypothesis.

```{r}
mex <- getData('GADM', country='MEX', level=1)
pm <- predict(wc, rrf, ext=mex)
pm <- mask(pm, mex)
plot(pm)
```

**Question 4:** Where in Mexico are you most likely to encounter western bigfoot?

According to the random forest model predictions, we are most likely to observe western Bigfoot in the northern tip of Baja California.

```{r}
# load future climate data estimates
fut <- getData('CMIP5', res=10, var='bio', rcp=85, model='AC', year=70)
# keep consistent variable names
names(fut) <- names(wc)
# make predictions for future climate scenario
futusa <- predict(fut, rrf, ext=eus, progress='window')

plot(futusa)
```

**Question 5:** Make a map to show where conditions are improving for western bigfoot, and where they are not. Is the species headed toward extinction?

```{r}
plot(futusa - rc2)
```

According to the random forest model predictions under the climate change scenario, it appears there will be improved habitat conditions in coastal British Columbiam but much of the coastal US and the Midwest will lose Bigfoot habitat. Overall, it appears that much more habitat will be lost than gained, suggesting potential negative consequences for western Bigfoot in the future.

